---
title: "Lab 3: Data Wrangling for Visualization"
author: "Joshua Liggins"
editor: visual
format:
  html:
    toc: true
    toc-depth: 3
    code-copy: true
    code-tools: true
execute:
  eval: false
  warning: false
  message: false
---
# Creating the Raw Data

## Load Packages 
```{r}
library(tidyverse)
```

## Raw-data
```{r}
raw_health_data <- tibble(
  county = c(
    "Clayton",
    "Cobb",
    "Dekalb",
    "Fulton",
    "Gwinnett",
    "Henry",
    "Rockdale",
    "Cherokee",
    "Douglas",
    "Fulton"
  ),
  region = c(
    "Central",
    "North",
    "Central",
    "Central",
    "North",
    "South",
    "Central",
    "North",
    "South",
    "Central"
  ),
  population_2023 = c(
    300000,
    770000,
    760000,
    1080000,
    970000,
    250000,
    95000,
    300000,
    150000,
    1080000
  ),
  cases_2023 = c(
    4200,
    5800,
    6100,
    8500,
    6900,
    3200,
    1800,
    2500,
    2100,
    8500
  ),
  population_2024 = c(
    305000,
    780000,
    765000,
    1090000,
    985000,
    255000,
    97000,
    310000,
    155000,
    1090000
  ),
  cases_2024 = c(
    4500,
    6000,
    NA,
    8800,
    7100,
    3400,
    1900,
    2700,
    2200,
    8800
  )
)
raw_health_data
```

# Diagnose

## Diagnose-data

```{r}
glimpse(raw_health_data)
```
```{r}
raw_health_data %>%
  count(county) %>%
  filter(n > 1)
```
```{r}
raw_health_data %>%
  summarize(
    across(
      everything(),
      ~ sum(
        is.na(.)
      )
    )
  )
```
# Remove Duplicates 

## removing dupicates

```{r}
health_data_clean <- raw_health_data %>%
  distinct(
    county,
    .keep_all = TRUE
  )
```

```{r}
health_data_clean %>%
  count(county)
```

# Reshape Long 

## Reshape-long 

```{r}
health_data_long <- health_data_clean %>%
  pivot_longer(
    cols = c(
      population_2023,
      cases_2023,
      population_2024,
      cases_2024
    ),
    names_to = c(
      ".value",
      "year"
    ),
    names_pattern = "(population|cases)_(\\d{4})"
  ) %>%
  mutate(
    year = as.integer(year)
  )
health_data_long
```

# Calculate Rates

## Calculate-rates 

```{r}
health_data_long <- health_data_long %>%
  mutate(
    rate_per_100000 = (cases / population) * 100000
  )
health_data_long
```

# Summarize 

## Summarize-region

```{r}
regional_summary <- health_data_long %>%
  group_by(
    region, 
    year
  ) %>%
  summarize(
    mean_rates = mean(
      rate_per_100000, 
      na.rm = TRUE
    ),
    .groups = "drop"
  )
regional_summary
```

# Join Contextual Data 

## Join data 

```{r}
rural_status <- tibble(
  county = c(
    "Clayton",
    "Cobb",
    "Dekalb",
    "Fulton",
    "Gwinnett",
    "Henry",
    "Rockdale",
    "Cherokee",
    "Douglas"
  ),
  rural_status = c(
    "Urban",
    "Urban",
    "Urban",
    "Urban",
    "Suburban",
    "Suburban",
    "Suburban",
    "Suburban",
    "Rurual"
  )
)
rural_status
```

```{r}
health_data_joined <- health_data_long %>%
  left_join(
    rural_status,
    by = "county"
  )
```

```{r}
rural_status %>%
  count(county) %>%
  filter( n > 1 )
```

```{r}
health_data_joined %>%
  filter(
    is.na(
      rural_status
    )
  )
```

```{r}
nrow(health_data_long)
nrow(health_data_joined)
```

# Visualize

## Visualizations 

### Regional line graph 

```{r}
regional_summary %>%
  ggplot(
    aes(
      x = year,
      y = mean_rates,
      group = region,
      color = region
    )
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(
    title = "Mean Health Case Rate by Region",
    x = "Year",
    y = "Cases per 100,000",
    color = "Region"
  ) +
  theme_minimal()
```

### Ranked 2024 county graph

```{r}
health_data_joined %>%
  filter(year == 2024) %>%
  ggplot(
    aes(
      x = rate_per_100000,
      y = reorder(county, rate_per_100000)
    )
  ) +
  geom_col() +
  labs(
    title = "County Health Case Rates in 2024",
    x = "Cases per 100,000",
    y = "County"
  ) +
  theme_minimal()
health_data_joined
```

### Rural-status comparison

```{r}
health_data_joined %>%
  filter(year == 2024) %>%
  ggplot(
    aes(
      x = rural_status,
      y = rate_per_100000,
      fill = rural_status
    )
  ) +
  geom_boxplot() +
  labs(
    title = "Health Case Rates by Rural Status in 2024",
    x = "Rural Status",
    y = "Cases per 100,000",
    fill = "Rural Status"
  ) +
  theme_minimal()
health_data_joined
```

# Reflection 

1. Why was long format useful? 
- Long format was useful becasue it organized the data so that each row reprsented a county-year combination. This made it easier to calculate rates, summarize the data by region and year, and create visualizations using `ggplot2`. It also allowed the same variables, such as population and cases, to be used consistently across mutiple years. 

2. What causes row multiplication during joins? 
- Row multiplication can occur when the dataset being joined contains multiple rows for the same key, such as a county. If a county appears more than once in the contextual dataset, `left_joint()` can match multiple rows to the same county in the health dataset. This can create duplicate observations and lead to incorrect summaries and visualizations. 

3. Why are rates preferable to counts? 
- Rates are preferable to counts when comparing populations of different sizes because they account for differences in population. A county with a larger population may have more cases simply becasue more people live there. Using cases per 100,000 allows for a more meaningful comparison of disease burden across counties and regions.  
















