---
title: "Lab 2 Applying the Grammar of Graphics"
author: "Joshua Liggins"
subtitle: "Student Version"
format:
  html:
    toc: true
    toc-depth: 3
    code-copy: true
    code-tools: true
execute:
  eval: false
  warning: false
  message: false
---

## Loading Packages

```{r}
library(tidyverse)
library(gapminder)
```

## Prepare the Data

```{r}
gapminder_2007 <- gapminder %>%
  filter(year == 2007) %>%
  rename(
    life_expectancy = lifeExp,
    gdp_per_capita = gdpPercap,
    population = pop
  )
```

## Scatterplot

```{r}
ggplot(gapminder_2007,
       aes(
         x = gdp_per_capita,
         y = life_expectancy,
         color = continent
       ))+
  geom_point(
    size = 3,
    alpha = 0.8) +
  scale_x_log10()+
  scale_color_viridis_d()+
  labs(
    title = "GDP per capita and Life Expectancy in 2007",
    x = "GDP per capita (US$)",
    y = "Life Expectancy (Years)",
    color = "Continent"
  )+
  theme_minimal()
```

## Bar-Chart

```{r}
# First I will calculate the mean life expectancy for each continent, then arrange the continents from lowest to highest.

life_expectancy_continent <- gapminder_2007 %>%
  group_by(continent) %>%
  summarize(
    mean_life_expectancy = mean(life_expectancy)
  ) %>%
  arrange(
    mean_life_expectancy
  )
ggplot(
  life_expectancy_continent,
  aes(
    x = mean_life_expectancy,
    y = reorder(
      continent,
      mean_life_expectancy
    )
  )
)+
  geom_col()+
  labs(
    title = "Mean Life Expectancy by Continent in 2007",
    x = "Mean Life Expectancy (Years)",
    y = "Continent"
  )+
  theme_classic()
```

## Boxplot

```{r}
ggplot(gapminder_2007,
       aes(
         x = continent,
         y = life_expectancy,
         fill = continent
       )) +
  geom_boxplot()+
  scale_fill_viridis_d() +
  labs(
    title = "Life Expectancy Across Continents in 2007",
    x = "Continent",
    y = "Life Expectancy (Years)",
    fill = "Continent"
  ) +
  theme_classic()
```

## Faceting

```{r}
ggplot(gapminder_2007,
       aes(
         x = gdp_per_capita,
         y = life_expectancy,
       ))+
  geom_point(
    size = 2.5, 
    alpha = 0.8
  )+
  scale_x_log10()+
  facet_wrap(~ continent) +
  labs(
    title = "GDP per Capita and Life Expectancy by Continent in 2007",
    x = "GDP per Capita (US$)",
    y = "Life Expectancy (Years)"
  )+
  theme_minimal()
```

## Identify the Grammar 

The scatterplot in part two of the lab uses the `gapminder_2007` as the dataset. The aesthetics map GDP per capita to the x-axis, life expectancy to the y-axis, and continent to the color. The geometry is `geom_point()`, which creates the scatterplot. The scale uses `scale_x_log10()` to display GDP per capita on a logarithmic scale and `scale_color_viridis_d()`for an accessible color palette. The labels include a title, descriptive x-axis and y-axis, and a legend title for continent. No facets are used in this graph. The theme is `theme_minimal()`which provides a clean and readable appearance. 

## Independent Chart 

```{r}
ggplot(gapminder_2007,
       aes(
         x = gdp_per_capita,
         y = life_expectancy,
         color = continent
       ))+
  geom_point(
    size = 3,
    alpha = 0.8
  )+
  geom_smooth(
    aes(group = 1),
    method = "lm",
    se = FALSE,
    color = "black"
  ) +
  scale_x_log10()+
  scale_color_viridis_d()+
  labs(
    title = "Economic Resources and Life Expectancy Across Countries in 2007",
    x = "GDP per Capita (US$)",
    y = "Life Expectancy (Years)",
    color = "Continent"
  ) +
  theme_minimal()
```

## Reflection

1. Which aesthetic mappings changed the meaning of the graph?

- Coloring the points by continent added geographic context to the graph and made it possible to compare patterns across different regions. Mapping GDP per capita to the x-axis and life expectancy to the y-axis emphasized the relationship between economic resources and population health. 

2. When did faceting improve interpretation?

- Faceting improved interpretation when examining the GDP-life expectancy relationship separately by continent. It made it easier to see patterns within each continent without having all countries represented in the same panel. 

3. Which graph most clearly communicated one finding?

- The independent scatterplot most clearly communicated the finding that countries with higher GDP per capita generally tend to have higher life expectancy. The logarithmic x-axis and trend line made the overall relationship easier to identify. 



